EdgeR analysis for differential gene expression
```{r}
#Loading necessary libraries and functions
source("~/MM_thesis_repo/libraries_thesis.R")
source("~/MM_thesis_repo/functions_MM.R")
```

Setting up the variables used in the script 
```{r}
analysis <- "MA_14_MM_ND_RNA"
analysis_title <- "expression in MM patients and normal donors"
Group <- c(rep("MM", 10), rep("ND", 5)) #change the group if more groups present 
# use Batch to define which samples are batch pairs (set all to 1 if no batch effect)
Batch <- 1
desired_contrast <- "Expression = groupND-groupMM" #put as a list to have multiple comparisons 

save_outputs <- TRUE  # Change to FALSE if you don't want to save outputs, TRUE if outputs are to be saved 
```


Reading the data
```{r}
counts <- read_tsv("~/myeloma_thesis_R/general_data/patients_feature_counts.txt", skip = 1) %>%
 set_names(c(
    "Geneid", "chr", "start", "end", "strand", "length", "CCND1_RS_3B.4", "CCND1_RS_1.4", "CCND1_RS_1.12", "CCND1_RS_1.19", "MMSET_RS_3B.3", "MMSET_RS_1.14", "MMSET_RS_2.6", "MMSET_RS_1.3", "MAF_RS_1.16", "MAF_RS_1.15", "ND_RS_1.22", "ND_RS_1.23", "ND_RS_2.8", "ND_RS_2.9", "ND_RS_4.25"
    ))
```

Set up analysis dataframe and design matrix
```{r}
target <- data.frame(
  # choose which samples (which columns) to analyse
  Samples = colnames(counts)[c(7:21)] %>% factor(),
  Group = Group,
  Batch = Batch
)

target

group <- target$Group
design <- model.matrix(~0+group)
design
```
Create list of DGEs
```{r}
counts_dge <- DGEList(counts = counts[,7:21], group=group, genes = counts[,1])
print(counts_dge$samples)

counts_cpm <- cpm(counts_dge)
```

Filter out enhancers with low tag in at least 5 samples 
```{r}
print("Genes before filtering:")
nrow(counts_dge$counts)

filter_cpm <- rowSums(counts_cpm > 1) >= 5 # Keep genes with > 1 cpm in at least the min num of patients within one group
counts_dge <- counts_dge[filter_cpm,]

print("Genes after filtering:")
nrow(counts_dge$counts)
```
```{r}
gene_list <- counts_dge$genes
```


Apply normalisation
```{r}
counts_norm <- calcNormFactors(counts_dge)
counts_norm$samples # Shows normalisation effect
```

Generate PCA plots to look at sample similarity (two alternative plotting scripts)
```{r}
mds <- plotMDS(counts_norm)

mds.df <- data.frame(
  MDS1 = mds$x,
  MDS2 = mds$y,
  Group = group,
  Sample = target$Samples
)

p.pca <- ggplot(
  mds.df, 
  aes(
    x = MDS1, 
    y= MDS2, 
    fill=Group, 
    label=Sample)
  ) +
  geom_point(
    pch=21, size=3
    ) +
  geom_text_repel() +
  ggtitle(
    paste0("MDS of ", analysis_title)
    ) +
  theme_classic()
p.pca

if (save_outputs) {
  ggsave(paste0(analysis, "_MDSplot.pdf"), p.pca,height=5,width=5, useDingbats=F)
}
```
Plot a distance matrix
```{r}
distsRL <- dist(t(counts_norm$counts))
mat <- counts_norm$counts %>%
  t() %>%
  dist() %>%
  as.matrix()

#rownames(mat) <-  group
#colnames(mat) <-  group

hmcol <- colorRampPalette(brewer.pal(9, "Blues"))(255)

if (save_outputs) {
  pdf(paste0(analysis, "_distancematrix.pdf"), 5, 5)
  heatmap(mat, trace="none", col = rev(hmcol))
  dev.off()
}

```

Estimate dispersion and model fitting
```{r}
set_d <- estimateDisp(counts_norm, design)
print("Common dispersion:")
print(set_d$common.dispersion)

pdf(paste0(analysis, "_BCVplot.pdf"), 5,5)
plotBCV(set_d)
dev.off()

fit <- glmFit(set_d, design)
```

Set up contrast groups
```{r}
contrasts <- makeContrasts(
  desired_contrast,
  levels = design
  )
lrt <- glmLRT(fit, contrast=contrasts[,1])
```

Extract DEGs and export: Group1 vs Group2
```{r}
lrt.top_tags <- topTags(
  lrt, n=nrow(counts_norm)
  )

lrt.table <- lrt.top_tags$table %>%
  dplyr::select(Geneid, logFC, logCPM, LR, PValue, FDR)

res.counts <- lrt.table %>%
  left_join(data.frame(Geneid = counts_norm$genes, counts_norm$counts))

res.cpm <- lrt.table %>%
  left_join(data.frame(Geneid = counts[,1], counts_cpm))

if (save_outputs) {
 write_tsv(res.counts, paste0(analysis, "_counts.txt"))
 write_tsv(res.cpm, paste0(analysis, "_cpm.txt"))
}
```

MA plot
```{r}
p.ma <- ggplot(mapping = aes(x=logCPM, y=logFC)) +
  geom_point(data = res.counts[res.counts$FDR >= 0.05,], col="black") +
  geom_point(data = res.counts[res.counts$FDR < 0.05,], col="red") +
  ggtitle(paste0("MA plot of ", analysis_title)) +
  theme_classic()
p.ma

if (save_outputs) {
ggsave(paste0(analysis, "_MAplot.pdf"), p.ma, height=5, width=5, useDingbats=F)
}
```
Volcano plot
```{r}
p.volcano <- res.counts %>%
  dplyr::mutate(class = case_when(
    (FDR <= 0.05 & abs(logFC) < 1) ~ "FDR<0.05",
    (FDR > 0.05 & abs(logFC) >= 1) ~ "logFC>1",
    (FDR <= 0.05 & abs(logFC) >= 1) ~ "Both",
    (FDR > 0.05 & abs(logFC) < 1) ~ "Unchanged"
  )) %>%
  ggplot(aes(x = logFC, y=-log10(PValue), col=class)) +
    geom_point() +
    scale_color_manual(values = c("FDR<0.05"="red2", "logFC>1"="orange", "Both"="green3", "Unchanged" = "grey60")) +
    ggtitle(paste0("Volcano plot of ", analysis_title)) +
    theme_classic()
p.volcano


if (save_outputs) {
ggsave(paste0(analysis, "_volcano.pdf"), p.volcano, height=5, width=5, useDingbats=F)
}
```
